---
title: "NHS Borders | Cancer Incidence Rates"
output:
  html_document:
    df_print: paged
    toc: TRUE
    toc_float: TRUE
---

```{r, eval = TRUE, echo = FALSE, message = FALSE}
# load in the tidyverse
library(tidyverse)
```

```{r, eval = TRUE, echo = FALSE, message = FALSE}
# read in cancer incidence data
cancer_data_by_HB <- read_csv("data/clean_data/cancer_data_by_HB_clean.csv")
# read in lookup tibble
HB_lookup_tibble <- read_csv("data/raw_data/geo_HB_labels_lookup.csv") %>% 
  janitor::clean_names()
```
## 1. Introduction and Aims

### Brief
In order to help inform the planning for provision of cancer treatment services
in NHS Borders, we would like to gain better understanding of the incidence of
cancer in NHS Borders.

### Investigation Aims

The aims of this investigation are to answer the following questions: 

1. What is the trend in cancer incidence over the period?
2. What types of cancer are the most common?
3. What types of cancer have the highest increase in incidence rate over the
period?
4. Which types of cancer are most prevalent among men?
5. Which types of cancer are most prevalent among women?

## 2. Analysis
```{r, eval = TRUE, echo = FALSE}
# get area code for NHS Borders from lookup tibble
borders_hb_code <- HB_lookup_tibble %>% 
  filter(hb_name == "NHS Borders") %>% 
  select(hb) %>% 
  pull()

# filter cancer data for only NHS Borders Health Board
cancer_data_borders <- cancer_data_by_HB %>%
  filter(hb == borders_hb_code)
```

```{r, eval = TRUE, echo = FALSE}
# get names of specific cancer sites (ignore "All")
specific_cancers <- cancer_data_borders %>%
  distinct(cancer_site) %>% 
  filter(!str_detect(cancer_site, "All")) %>% 
  pull()
```

```{r, eval = TRUE, echo = FALSE}
# create function that filters the data and calculates the sum of all cancer
# incidences for each cancer site
generate_sum_incidences_per_cancer_site <-
  function(data, sex = "All", year = 1994) {
  data %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(measurement == "CRUDE") %>% 
  filter(sex == sex) %>% 
  filter(year >= year) %>% 
  group_by(cancer_site) %>% 
  summarise(total_incidences = sum(incidences_all_ages), .groups = "drop_last")
  }
```

```{r, eval = TRUE, echo = FALSE}
sum_incidences_per_cancer_site <- generate_sum_incidences_per_cancer_site(
  cancer_data_borders
)
```

### 1. What is the trend in cancer incidence over the period (1994 - 2018)?

One of the cancer sites listed in the data was "All cancer types". By filtering
for this condition an overview of all cancer incidences for all sexes can be
plotted:

```{r, eval = TRUE, echo = FALSE}
cancer_data_borders %>% 
  filter(cancer_site == "All cancer types") %>% 
  filter(sex == "All") %>% 
  ggplot() +
  aes(
    x = year,
    y = incidences_all_ages
  ) +
  geom_line() +
  facet_wrap(~sex) +
  scale_x_continuous(breaks = c(1995,2000,2005,2010,2015)) +
  theme_bw() +
  labs(x = "Year",
       y = "number of Cancer Incidences",
       title = "Number of Cancer Incidences vs Year")
```

One of the variables was sex, so the same plot could be split into the two sexes
for a more specific picture:

```{r, eval = TRUE, echo = FALSE}
cancer_data_borders %>% 
  filter(cancer_site == "All cancer types") %>% 
  filter(sex != "All") %>% 
  filter(measurement == "CRUDE") %>% 
  ggplot() +
  aes(
    x = year,
    y = incidences_all_ages
  ) +
  geom_line() +
  facet_wrap(~sex, ncol = 1) +
  scale_x_continuous(breaks = c(1995,2000,2005,2010,2015)) +
  theme_bw() +
  labs(x = "Year",
       y = "number of Cancer Incidences",
       title = "Number of Cancer Incidences vs Year")
```

Using total cancer incidences for each year provides a general overview. A more
specific overview could use the crude rate, which is a measure which takes the
population into account.

```{r, eval = TRUE, echo = FALSE}
cancer_data_borders %>% 
  filter(cancer_site == "All cancer types") %>% 
  filter(sex != "All") %>% 
  filter(measurement == "CRUDE") %>%
  ggplot() +
  aes(
    x = year,
    y = value
  ) +
  geom_line() +
  facet_wrap(~sex, ncol = 1) +
  scale_x_continuous(breaks = c(1995,2000,2005,2010,2015)) +
  theme_bw() +
  labs(x = "Year",
       y = "Crude Rate Value",
       title = "Number of Cancer Incidences vs Year")
```

If it were the case that cancer incidence was increasing due to the nature of an
increase in population alone, this line would be flat.

### 2. Which types of cancer are the most common?

After filtering the data, this table was generated. It shows the cancer site (
type of cancer) and the number of recorded incidences for the period (1994 - 
2018)

```{r, eval = TRUE, echo = FALSE}
generate_sum_incidences_per_cancer_site <- function(data, sex = "All", year = 1994) {
  data %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(measurement == "CRUDE") %>% 
  filter(sex == sex) %>% 
  filter(year >= year) %>% 
  group_by(cancer_site) %>% 
  summarise(total_incidences = sum(incidences_all_ages), .groups = "drop_last")
  }
```


```{r, eval = TRUE, echo = FALSE}
sum_incidences_per_cancer_site <- generate_sum_incidences_per_cancer_site(cancer_data_borders) %>% 
  slice_max(total_incidences, n = 10) %>% 
  arrange(desc(total_incidences)) %>% 
  rename(`Cancer Site` = cancer_site,
         `Total Incidences` = total_incidences)

sum_incidences_per_cancer_site
```

```{r, eval = TRUE, echo = FALSE}
make_bar_plot <- function(data) {
  data %>% 
    ggplot() + 
    aes(x = `Cancer Site`,
        y = `Total Incidences`) +
    geom_col() +
    theme_bw() + 
    theme(axis.text.y = element_text(angle = 0, hjust = 1)) +
    coord_flip() +
  labs(x = "Cancer Type",
       y = "Total n of Incidences")
}
```

The total number of incidences for each type of cancer recorded in the Borders
can be represented graphically:

```{r, eval = TRUE, echo = FALSE}
sum_incidences_per_cancer_site %>% 
  make_bar_plot() 
```

For a more in depth understanding of the most common cancer types, the ten with
the highest number of total incidences over the period (1994 - 2018) can be
plotted:

```{r, eval = TRUE, echo = FALSE}
sum_incidences_per_cancer_site %>% 
  slice_max(`Total Incidences`, n = 10) %>% 
  make_bar_plot() +
  ggtitle("10 Most Common Cancer Types")
```

To understand if there has been any change in the past decade, the data can be
filtered and then replotted.

This shows the 10 most common cancer types in the past decade only.

```{r, eval = FALSE, echo = FALSE}
sum_incidences_per_cancer_site_recent <- generate_sum_incidences_per_cancer_site(
  cancer_data_borders,
  year = "2009"
)
```

```{r, eval = FALSE, echo = FALSE}
sum_incidences_per_cancer_site_recent %>% 
  rename(`Cancer Site` = cancer_site,
         `Total Incidences` = total_incidences) %>% 
  slice_max(`Total Incidences`, n = 10) %>% 
  make_bar_plot()+
  ggtitle("10 Most Common Cancer Types (since 2009)")
```

### 3. Which types of cancer have the highest increase in incidence rate over the period (1994 - 2018)?

Examination of crude rates:

```{r, eval = TRUE, echo = FALSE}
cancer_data_borders %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(sex == "Male") %>% 
  filter(measurement == "CRUDE") %>%
  group_by(cancer_site) %>% 
  summarise(year, diff_crude_rate = value - lag(value),
            .groups = "drop_last") %>% 
  group_by(cancer_site) %>% 
  summarise(mean_diff = mean(diff_crude_rate, na.rm = TRUE),
            .groups = "drop_last") %>% 
  slice_max(mean_diff, n = 10) %>%
  ggplot() +
  aes(x = cancer_site,
      y = mean_diff) + 
  geom_col() +
  coord_flip() +
  theme_bw() +
  
  labs(x = "Cancer Type",
       y = "Average yearly change",
       title = "Mean increase in crude rate | Male") +
  theme(plot.title = element_text(hjust = 0.5))
  
```

```{r, eval = TRUE, echo = FALSE}
cancer_data_borders %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(sex == "Female") %>% 
  filter(measurement == "CRUDE") %>%
  group_by(cancer_site) %>% 
  summarise(year, diff_crude_rate = value - lag(value),
            .groups = "drop_last") %>% 
  group_by(cancer_site) %>% 
  summarise(mean_diff = mean(diff_crude_rate, na.rm = TRUE),
            .groups = "drop_last") %>% 
  slice_max(mean_diff, n = 10) %>%
  ggplot() +
  aes(x = cancer_site,
      y = mean_diff) + 
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(x = "Cancer Type",
       y = "Average yearly change",
       title = "Mean increase in crude rate | Female") +
  theme(plot.title = element_text(hjust = 0.5))
```


### 4. Which types of cancer are most prevalent among men?

```{r, eval = TRUE, echo = FALSE}
highest_cancers_2018_male <- cancer_data_borders %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(measurement == "CRUDE") %>% 
  filter(sex == "Male") %>% 
  filter(year == "2018") %>% 
  group_by(cancer_site) %>% 
  summarise(mean_c_rate = mean(value, na.rm = TRUE), .groups = "drop_last") %>% 
  slice_max(mean_c_rate, n = 5) %>% 
  select(cancer_site) %>% 
  pull()
```


```{r, eval = TRUE, echo = FALSE}
cancer_data_borders %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(cancer_site %in% highest_cancers_2018_male) %>% 
  filter(measurement == "CRUDE") %>% 
  filter(sex == "Male") %>%
  group_by(cancer_site) %>%
  ggplot() +
  aes(x = year, y = value, group = cancer_site, colour = cancer_site) +
  geom_line() +
  labs(x = "Year",
       y = "Crude Rate",
       title = "Cancer types common among men") +
  facet_wrap(~cancer_site) +
  scale_x_continuous(breaks = c(1995,2000,2005,2010,2015)) +
  theme_bw() +
  theme(legend.position = "none")
```


### 5. Which types of cancer are most prevalent among women?

```{r, eval = TRUE, echo = FALSE}
highest_cancers_2018_female <- cancer_data_borders %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(measurement == "CRUDE") %>% 
  filter(sex == "Female") %>% 
  filter(year == "2018") %>% 
  group_by(cancer_site) %>% 
  summarise(mean_c_rate = mean(value, na.rm = TRUE), .groups = "drop_last") %>% 
  slice_max(mean_c_rate, n = 5) %>% 
  select(cancer_site) %>% 
  pull()
```


```{r, eval = TRUE, echo = FALSE}
cancer_data_borders %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(cancer_site %in% highest_cancers_2018_female) %>% 
  filter(measurement == "CRUDE") %>% 
  filter(sex == "Female") %>%
  group_by(cancer_site) %>%
  ggplot() +
  aes(x = year, y = value, group = cancer_site, colour = cancer_site) +
  geom_line() +
  facet_wrap(~cancer_site) +
  scale_x_continuous(breaks = c(1995,2000,2005,2010,2015)) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "Year",
       y = "Crude Rate",
       title = "Cancer types common among women")
```

### Combined Plot

```{r, eval = TRUE, echo = FALSE}
highest_cancers_2018_both <- cancer_data_borders %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(measurement == "CRUDE") %>% 
  filter(sex != "All") %>% 
  filter(year == "2018") %>% 
  group_by(cancer_site) %>% 
  summarise(mean_c_rate = mean(value, na.rm = TRUE), .groups = "drop_last") %>% 
  slice_max(mean_c_rate, n = 5) %>% 
  select(cancer_site) %>% 
  pull()
```


```{r, eval = TRUE, echo = FALSE}
cancer_data_borders %>% 
  filter(cancer_site %in% specific_cancers) %>% 
  filter(cancer_site %in% highest_cancers_2018_both) %>% 
  filter(measurement == "CRUDE") %>% 
  filter(sex != "All") %>%
  group_by(cancer_site) %>%
  ggplot() +
  aes(x = year, y = value, group = cancer_site, colour = cancer_site) +
  geom_line() +
  labs(x = "Year",
       y = "Crude Rate",
       title = "Trends in crude rate over time for common cancers for each sex") +
  facet_grid(vars(sex), vars(cancer_site)) +
  theme_bw() +
  theme(legend.position = "none")
```



```{r}

```



